---
title: 'PCA, FA and Clustering - Project in Statistics'
output:
  html_document:
    df_print: paged
---

## Learning outcomes

In this notebook we will find out what it's like to "play" with big data tables with a lot of variables, cluster them using the k-means process and perform Principal Component and Factor Analysis.

Also, we will see how to create beautiful graphs and draw heated maps.

## Install required packages

```{r eval=FALSE}
install.packages("tidyverse")
install.packages("mosaic")
install.packages("ggplot2")
install.packages("moments")
install.packages("sjlabelled")
install.packages("Hmisc")
install.packages("googledrive")
```

## Load required packages

```{r loadlib, echo=T, results='hide', message=F, warning=F}
library(tidyverse)
library(GGally)
library(dplyr)
library(factoextra)
library(rworldmap)
library(countrycode)
library(googledrive)
```


## Import the dataset. Remeber it's better to have your data in .csv format!

```{r eval=FALSE}
dat <- read.csv("WBD 1.csv", header = TRUE)
View(dat)
# We want to see how many rows our dataset has
nrow(dat)
```


## Given the fact that we are using the whole world as datasample we cannot hace more than 217 rows as there are 217 countries in WB database.

```{r loadlib, echo=T, results='hide', message=F, warning=F}
## For example we can see below that on row 218, all columns included, the country is "Arab World"
dat.218 <- dat[218, ]
dat.218
## A second example is the row 250, all columns included, where it corresponds to the "North America" region
dat.250 <- dat[250, ]
dat.250
```


## Since we have regions apart from countries we have to eliminate rows 218 until 269 #

```{r loadlib, echo=T, results='hide', message=F, warning=F}
## First we create a subset of the last 52 rows and sum them to see what are the regions that are not single countries that have to be deleted.
dat.last52 <- dat[218:269, ]
str(dat.last52)
### Now we can remove the separated dataframe from the original dataset and obtain our desired dataframe to work with
dat.remove.last52 <- dat[-c(218:269), ]

```


## Rename the new dataframe. This is now the dataset we will work with after clearing the missing values

```{r eval=FALSE}
WBD <- dat.remove.last52[1:217, ]
```


## Are there any Missing Values?

```{r eval=FALSE}
# Since our mising values are denoted as ".." we have to rename them as "NA" in order to eliminate them.
WBD[WBD==".."] = NA

# We test to see if we have any missing values, now that we have changed the notation properly.
is.na(WBD)

# It is useful to see how many NA's we have in our dataframe. 
sum(is.na(WBD[]))

# We can observe that all of the rows contain at least on missing "NA" value. Thus the option complete.cases is not useful as we will have 0 observations after 
# the cleaning process.
# With colSums we can see how many missing variables we have on each column and decide how to eliminate them according to the weight of our variables.
colSums(is.na(WBD))
```


## Dealing with the missing values. Elimination Process

```{r eval=FALSE}
#First we see how many rows we have now so to compare how much our dataset will shrink
nrow(WBD)

# We create a dataframe with all the rows and transfer across columns that have less than 100 missing values
WBD.clean1 <- WBD[ ,colSums(is.na(WBD))<100]
# Then we strip out the NA's and summarize the final sample we will work with
WBD.clean.final <- na.omit(WBD.clean1)
# Our dataframe slightly drops in observations but this is completely rational given the fact that the development indexes we chose are not available for the African Continent.
nrow(WBD.clean.final)

# We test to see if there are still any missing data in our dataframe. It turns out a logical var (TRUE or FALSE) to indicate the presence of "NA's".
any(is.na(WBD.clean.final[]))

str(WBD.clean.final)
```


## We  rename our variables to plot easier and provie more clear information

```{r eval=FALSE}
WBD.final <- WBD.clean.final %>% rename("Year" = Time, "Country"=`Country Name`, "Female Employers (% of female employment)" = `Employers, female (% of female employment) (modeled ILO estimate) [SL.EMP.MPYR.FE.ZS]`, "Male Employers (% of male employment)" = `Employers, male (% of male employment) (modeled ILO estimate) [SL.EMP.MPYR.MA.ZS]`, "Employment in industry, male (% of male employment)" = `Employment in industry, male (% of male employment) (modeled ILO estimate) [SL.IND.EMPL.MA.ZS]` , "Employment in industry, female (% of female employment)" = `Employment in industry, female (% of female employment) (modeled ILO estimate) [SL.IND.EMPL.FE.ZS]` , "GDP (current US$)" = `GDP (current US$) [NY.GDP.MKTP.CD]` , "GDP per capita (current US$)" = `GDP per capita (current US$) [NY.GDP.PCAP.CD]` , "GDP per capita growth (annual %)" = `GDP per capita growth (annual %) [NY.GDP.PCAP.KD.ZG]` , "Population ages 65 and above (total)" = `Population ages 65 and above, total [SP.POP.65UP.TO]` , "Population ages 65 and above (% of total population)" = `Population ages 65 and above (% of total population) [SP.POP.65UP.TO.ZS]` , "Population ages 65 and above, female (% of female population)" = `Population ages 65 and above, female (% of female population) [SP.POP.65UP.FE.ZS]` , "Population ages 65 and above, male (% of male population)" = `Population ages 65 and above, male (% of male population) [SP.POP.65UP.MA.ZS]` , "Population growth (annual %)" = `Population growth (annual %) [SP.POP.GROW]` , "Start-up procedures to register a business (number)" = `Start-up procedures to register a business (number) [IC.REG.PROC]` , "Urban population growth (annual %)" = `Urban population growth (annual %) [SP.URB.GROW]` , "Urban population (% of total population)" = `Urban population (% of total population) [SP.URB.TOTL.IN.ZS]`) 

```


## Basic Descriptive statistics

```{r loadlib, echo=T, results='hide', message=F, warning=F}
dim(WBD.final)
str(WBD.final)

# A more detailed summary option 
glimpse(WBD.final)
```


## Our variables are depicted as factors, thus we convert them all into numericals.

```{r eval=FALSE}
WBD.final$Year <- as.numeric(WBD.final$Year)
WBD.final$`Male Employers (% of male employment)` <- as.numeric(WBD.final$`Male Employers (% of male employment)`)
WBD.final$`Employment in industry, male (% of male employment)` <- as.numeric(WBD.final$`Employment in industry, male (% of male employment)`)
WBD.final$`Employment in industry, female (% of female employment)` <- as.numeric(WBD.final$`Employment in industry, female (% of female employment)`)
WBD.final$Country <- as.character(WBD.final$Country)
WBD.final$`Female Employers (% of female employment)` <- as.numeric(WBD.final$`Female Employers (% of female employment)`)
WBD.final$`GDP (current US$)` <- as.numeric(WBD.final$`GDP (current US$)`)
WBD.final$`GDP per capita (current US$)` <- as.numeric(WBD.final$`GDP per capita (current US$)`)
WBD.final$`GDP per capita growth (annual %)` <- as.numeric(WBD.final$`GDP per capita growth (annual %)`)
WBD.final$`Population ages 65 and above (% of total population)` <- as.numeric(WBD.final$`Population ages 65 and above (% of total population)`)
WBD.final$`Population ages 65 and above, female (% of female population)` <- as.numeric(WBD.final$`Population ages 65 and above, female (% of female population)`)
WBD.final$`Population ages 65 and above, male (% of male population)` <- as.numeric(WBD.final$`Population ages 65 and above, male (% of male population)`)
WBD.final$`Population ages 65 and above (total)` <- as.numeric(WBD.final$`Population ages 65 and above (total)`)
WBD.final$`Population growth (annual %)` <- as.numeric(WBD.final$`Population growth (annual %)`)
WBD.final$`Start-up procedures to register a business (number)` <- as.numeric(WBD.final$`Start-up procedures to register a business (number)`)
WBD.final$`Urban population growth (annual %)` <- as.numeric(WBD.final$`Urban population growth (annual %)`)
WBD.final$`Urban population (% of total population)` <- as.numeric(WBD.final$`Urban population (% of total population)`)

str(WBD.final)
```


## Construct logs for large numbers such as GDP and Pop over 65

```{r loadlib, echo=T, results='hide', message=F, warning=F}
WBD.finalm = WBD.final
WBD.finalm[,10] = log(WBD.final[,10]) #GDP {log}
WBD.finalm[,11] = log(WBD.final[,11]) #GDP {log}
WBD.finalm[,13] = log(WBD.final[,13]) #Pop over 65 {log}
```


## Basic Descriptive statistics

```{r loadlib, echo=T, results='hide', message=F, warning=F}
dim(WBD.final)
str(WBD.final)
```


## A more detailed summary option

```{r loadlib, echo=T, results='hide', message=F, warning=F}
glimpse(WBD.final)
```


####################################################################################################
###################################### Plots, Pies and Graphs ######################################
####################################################################################################


## Basic Histogram 

```{r loadlib, echo=T, results='hide', message=F, warning=F}
plot(WBD.final$`Female Employers (% of female employment)`, WBD.final$`GDP per capita (current US$)`, xlab = "Female Empl", ylab = "GDPPC", frame.plot = FALSE, pch = 19, col = "blue"  )
plot(WBD.finalm$`GDP (current US$)`, WBD.finalm$`Start-up procedures to register a business (number)`, xlab = "Female Employers", ylab = "Start-up procedures to register a business (number)", frame.plot = FALSE, pch = 19, col = "blue"  )
```


## Basic Boxplot

```{r loadlib, echo=T, results='hide', message=F, warning=F}
boxplot(WBD.finalm[,6:20], las=2, col=rainbow(20))
```



## We plot GDPPC with urban population as a % of the total population of a country to see the correlation between the 2 variables


```{r loadlib, echo=T, results='hide', message=F, warning=F}
ggplot(WBD.finalm, aes(x=`GDP per capita (current US$)`, y= `Urban population (% of total population)`)) + geom_point() + labs(title="Countries", x="GDP per capita (current US$)", y="Urban population (% of total population)") + theme(legend.position="bottom")
```


## Basic Boxplot

```{r loadlib, echo=T, results='hide', message=F, warning=F}
boxplot(WBD.finalm[,6:20], las=2, col=rainbow(20))
```


## Plot with GDPPC and Urban Population:

```{r loadlib, echo=T, results='hide', message=F, warning=F}
ggplot(WBD.finalm, aes(x = `GDP per capita (current US$)`, y = `Urban population (% of total population)`, color = Region, size=`GDP (current US$)`/1e5)) + geom_point() + 
  scale_color_brewer(palette="Dark2")+ xlab("GDP per capita (current US$)") + ylab("Urban population (% of total population)")+theme_minimal()+ theme(legend.position="bottom")
```


## Other Plot with Female Employers (% of female employment) and Start-up procedures to register a business (number):

```{r loadlib, echo=T, results='hide', message=F, warning=F}
ggplot(WBD.finalm, aes(x = `Female Employers (% of female employment)`, y = `Start-up procedures to register a business (number)`, color = Region, size=`GDP (current US$)`/1e5)) + geom_point() + 
  scale_color_brewer(palette="BrBG")+ xlab("GDP per capita (current US$)") + ylab("Urban population (% of total population)")+theme_minimal()+ theme(legend.position="bottom")
```


## GDPPC and Urban Pop plots:

```{r loadlib, echo=T, results='hide', message=F, warning=F}
ggplot(WBD.finalm, aes(`GDP per capita (current US$)`)) + geom_density(aes(group=Region, colour=Region, fill=Region), alpha=0.1) +xlab("GDP per capita (current US$)")
ggplot(WBD.finalm, aes(`Urban population (% of total population)`)) + geom_density(aes(group=Region, colour=Region, fill=Region), alpha=0.1) +xlab("Urban population (% of total population)")
```


## Pop>65 plot:

```{r loadlib, echo=T, results='hide', message=F, warning=F}
ggplot(WBD.finalm, aes(`Population ages 65 and above (total)`)) + geom_density(aes(group=Region, colour=Region, fill=Region), alpha=0.1) +xlab("Population ages 65 and above (total)")
```


## Male Empl plot:

```{r loadlib, echo=T, results='hide', message=F, warning=F}
ggplot(WBD.finalm, aes(`Male Employers (% of male employment)`)) + geom_density(aes(group=Region, colour=Region, fill=Region), alpha=0.1) +xlab("Male Employers (% of male employment)") 
```


## Other relationship: Pop over 65 and GDPPC

```{r loadlib, echo=T, results='hide', message=F, warning=F}
ggplot(WBD.finalm, aes(x = `Population ages 65 and above (total)`, y = `GDP per capita (current US$)`, color=Region, label=Country)) + geom_text(size=2, check_overlap = TRUE) +
  theme_bw() + theme(legend.position="bottom")
```


## Other relationship: Pop over 65 and GDP

```{r loadlib, echo=T, results='hide', message=F, warning=F}

```


## Other relationship: Male and Female Employers

```{r loadlib, echo=T, results='hide', message=F, warning=F}
ggplot(WBD.finalm, aes(x = `Male Employers (% of male employment)`, y = `Female Employers (% of female employment)`, color=Region, label=Country)) + geom_text(size=2, check_overlap = TRUE) +
  theme_bw() + theme(legend.position="bottom")
```


## Plot separated according the regions

```{r loadlib, echo=T, results='hide', message=F, warning=F}
ggplot(WBD.finalm, aes(x = `GDP per capita (current US$)`, y = `Urban population (% of total population)`, color=`Population growth (annual %)`,size=`GDP (current US$)`)) +
  geom_point() + facet_wrap(~ Region) + scale_color_gradient(low="red", high="blue")+labs(title="Countries by region", x="GDP per Capita", y="Urban Population", colour="Annual Pop Growth (%)")+theme_minimal()+ theme(legend.position="bottom")
```


## Relations in dimension 2: bivariate analysis using scatter plots; multiple scatter plot: partial and all relations in dimension 2

```{r loadlib, echo=T, results='hide', message=F, warning=F}
pairs(WBD.finalm[,6:13],pch=19,col=WBD.finalm$Region)
pairs(WBD.finalm[,13:20],pch=19,col=WBD.finalm$Region)
pairs(WBD.finalm[,6:20],pch=19,col=WBD.finalm$Region)
```


## Correlation Graph

```{r loadlib, echo=T, results='hide', message=F, warning=F}
ggcorr(WBD.finalm, label = T, palette = "BrBG")+labs(title = "Correlations")
```
